---
title: "Copy number plots"
output: html_notebook
---

This notebook generates Figure 2B-D, as well as Supplementary Figures 1 and 2.

# Setup
```{r}
library(tidyverse)
library(ggpubr)
library(rstatix)
```

## Load data
```{r}
sample_sheet_paired <- read_delim('../results/2024-02-05_cohort-sheets/paired_sample_cohort_sheet.txt',
                                  show_col_types = FALSE)

sample_sheet_unpaired <- read_delim('../results/2024-02-05_cohort-sheets/unpaired_sample_cohort_sheet.txt',
                                  show_col_types = FALSE)

patient_sheet_paired <- read_delim('../results/2024-02-05_cohort-sheets/paired_participant_cohort_sheet.txt',
                                   show_col_types = FALSE)

patient_sheet_unpaired <- read_delim('../results/2024-02-05_cohort-sheets/unpaired_participant_cohort_sheet.txt',
                                     show_col_types = FALSE)

cnas_paired <- read_delim('../results/2024-01-18_gene-cna-calls/cnas_paired_all_annot_ptlevel.tsv',
                          show_col_types = FALSE)

cnas_unpaired <- read_delim('../results/2023-09-28_gene-cna-calls/cnas_unpaired_all_annot_ptlevel.tsv',
                            show_col_types = FALSE)

maf_paired <- read_delim('../results/2024-02-14_mutations/all_mut_ccfs_maf_bl_geneLists_evidenceLevels.txt',
                         show_col_types = FALSE)

maf_unpaired <- read_delim('../results/2024-02-12_mutations/all_mut_ccfs_maf_annotated_geneLists_evidenceLevels_unpaired.txt',
                         show_col_types = FALSE)

esr1_mut_pts <- maf_paired %>%
  filter(!str_equal(trajectory, 'low_ccf')) %>% 
  filter(str_equal(Hugo_Symbol, 'ESR1') &
         !str_equal(variant_type, 'silent')) %>% 
  select(Patient_ID, Hugo_Symbol) %>% 
  distinct %>%
  pull(Patient_ID)

esr1_mut_pts_unpaired <- maf_unpaired %>%
  filter(str_equal(Hugo_Symbol, 'ESR1') &
         !str_equal(variant_type, 'silent')) %>% 
  select(Patient_ID, Hugo_Symbol) %>% 
  distinct %>%
  pull(Patient_ID)

rm(maf_paired, maf_unpaired)
```

# Tidy data
```{r}
erbb2_tmp <- cnas_paired %>%
  filter(str_equal(gene, 'ERBB2')) %>% 
  separate_rows(sample, collection_date_dfd, CNAP, cna, sep = ';') %>% 
  mutate(CNAP = as.numeric(CNAP)) %>% 
  left_join(sample_sheet_paired %>% select(participant_id, collaborator_sample_id, pre_post, wxs_purity),
            by = c('sample' = 'collaborator_sample_id',
                   'participant_id' = 'participant_id'))

erbb2_tmp_per_pt <- erbb2_tmp %>% 
  group_by(participant_id, pre_post) %>%
  summarize(CNAP = mean(CNAP)) %>%
  pivot_wider(names_from = pre_post, values_from = CNAP)
```

# Combine data from paired and unpaired cohorts
```{r}
erbb2_unpaired_tmp <- cnas_unpaired %>% 
  filter(str_equal(gene, 'ERBB2')) %>% 
  separate_rows(CNAP, sample, sep = ';', convert = TRUE) %>% 
  left_join(sample_sheet_unpaired %>% select(collaborator_sample_id, submitted_material_type),
            by = c('sample' = 'collaborator_sample_id')) %>% 
  group_by(participant_id) %>% 
  summarize(CNAP = mean(CNAP)) %>% 
  left_join(patient_sheet_unpaired, by = 'participant_id')

esr1_mut_erbb2_cnap_combined <- bind_rows(
erbb2_unpaired_tmp,
erbb2_tmp_per_pt %>%
  drop_na %>% 
  left_join(patient_sheet_paired, by = 'participant_id') %>% 
  pivot_longer(cols = c(Pre, Post), names_to = 'Cohort', values_to = 'CNAP')
) %>% 
  ungroup %>% 
  select(participant_id, Cohort, CNAP, hr_status) %>% 
  mutate(esr1_category = factor(case_when(
  participant_id %in% c(esr1_mut_pts, esr1_mut_pts_unpaired) ~ 'Mutant',
  str_equal(hr_status, 'HR+') ~ 'HR+ wt',
  str_equal(hr_status, 'HR-') ~ 'HR- wt',
  TRUE ~ NA), levels = c('HR- wt', 'HR+ wt', 'Mutant')),
        group = factor(ifelse(Cohort %in% c('Pre', 'Non-exposed'), 'Pre-TKI', 'Post-TKI'),
                       levels = c('Pre-TKI', 'Post-TKI')),
        paired_or_unpaired = ifelse(Cohort %in% c('Pre', 'Post'), 'Paired', 'Unpaired'))
```

# Figure 2B: ERBB2 CNAP from pre- to post-TKI timepoints
```{r}
paired_cnap_plot <- ggpaired(erbb2_tmp_per_pt %>% drop_na, cond1 = 'Pre', cond2 = 'Post',
         line.color = 'gray', line.size = 0.5, xlab = 'Timing', point.size = 2) +
  stat_compare_means(label.y = 100, paired = TRUE, size = 4) +
  scale_y_continuous(limits = c(-10,110)) +
  theme_classic(base_size = 16)

paired_cnap_plot <- ggpar(paired_cnap_plot, ylab = bquote(italic('ERBB2')~'CNAP'))

paired_cnap_plot

# ggsave(paired_cnap_plot, filename = '../../results/2024-03-22_cna-plots/ERBB2_cnap_paired.png',
#        width = 3, height = 3, units = 'in', dpi = 450)
```

Does ERBB2 CNAP for a sample correlate with that sample's purity/tumor fraction?
```{r}
ggscatter(erbb2_tmp, x = 'wxs_purity', y = 'CNAP', add = 'reg.line', conf.int = TRUE,
          cor.coef = TRUE, cor.method = 'spearman')

cor.test(formula = ~ CNAP + wxs_purity, data = erbb2_tmp, method = 'spearman')
```

# Figure 2C: Pre-TKI ERBB2 CNAP by resistance type
```{r}
resistance_type_cnap_plot <- erbb2_tmp_per_pt %>%
  select(participant_id, Pre) %>% 
  drop_na %>% 
  left_join(patient_sheet_paired, by = 'participant_id') %>% 
  ungroup %>% 
  ggboxplot(., x = 'resistance_type', y = 'Pre',
            line.size = 0.5, add = 'jitter',
            add.params = list(size = 2),
            xlab = 'Resistance type',
            outlier.shape = NA) +
  theme_classic(base_size = 16) +
  stat_compare_means(method = 'wilcox.test', size = 4, label.y = 70)

resistance_type_cnap_plot <- ggpar(resistance_type_cnap_plot,
                                   ylab = bquote('Pre-TKI'~italic('ERBB2')~'CNAP'))

resistance_type_cnap_plot

# ggsave(resistance_type_cnap_plot, filename = '../../results/2024-03-22_cna-plots/ERBB2_cnap_resistance_type_pre.png',
#        width = 4, height = 4.5, units = 'in', dpi = 450)
```

# Figure 2D: ERBB2 CNAP by HR status and ESR1 mutation status
(combining the paired and unpaired cohorts)
```{r}
hrstatus_esrmut_combined_plot <- ggboxplot(esr1_mut_erbb2_cnap_combined %>% 
                                             filter(!is.na(esr1_category)), x = 'esr1_category', y = 'CNAP',
            palette = 'npg',
            facet.by = 'group',
            line.color = 'gray', line.size = 0.5,
            outlier.shape = NA) +
  geom_point(aes(shape = paired_or_unpaired),
             position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.9),
             size = 2) +
  stat_compare_means(method = 'wilcox.test',
                     comparisons = list(c('HR+ wt', 'HR- wt'),
                                        c('HR+ wt', 'Mutant'),
                                        c('HR- wt', 'Mutant'))) +
  theme_classic(base_size = 16) +
  theme(strip.background = element_blank(),
        legend.position = c(0.9, 0.5))

hrstatus_esrmut_combined_plot <- ggpar(hrstatus_esrmut_combined_plot,
                                       legend.title = 'Cohort',
                                       xlab = bquote(italic('ESR1')~'status'),
                                       ylab = bquote(italic('ERBB2')~'CNAP'))

hrstatus_esrmut_combined_plot$layers[[3]]$aes_params$textsize <- 4

hrstatus_esrmut_combined_plot

# ggsave(hrstatus_esrmut_combined_plot, filename = '../../results/2024-03-22_cna-plots/ERBB2_cnap_paired_and_unpaired_by_ESR1_and_HR.png',
#        width = 6, height = 5, units = 'in', dpi = 450)
```

# Figure 2B-D, combined
```{r}
manuscript_panels <- ggarrange(paired_cnap_plot,
                         resistance_type_cnap_plot,
                         hrstatus_esrmut_combined_plot,
                         nrow = 1,
                         widths = c(1,1,2))

manuscript_panels

# ggsave(manuscript_panels, filename = '../../results/2024-03-22_cna-plots/Fig2_BCD.pdf',
#        width = 14, height = 4.2, units = 'in')
```

# Supplementary Figure 1: ERBB2 CNAP by HR status
```{r}
# Reference: Pre vs. post, HR- vs. HR+
hrstatus_plot <- erbb2_tmp_per_pt %>% 
  drop_na %>% 
  left_join(patient_sheet_paired, by = 'participant_id') %>% 
  pivot_longer(cols = c(Pre, Post)) %>% 
  mutate(name = factor(name, levels = c('Pre', 'Post')),
         hr_status = factor(hr_status, levels = c('HR-', 'HR+'))) %>%
  ggboxplot(., x = 'hr_status', y = 'value',
            facet.by = 'name',
            line.color = 'gray', line.size = 0.5, ylab = 'ERBB2 CNAP',
            xlab = 'HR status', add = 'jitter') +
  stat_compare_means(label.y = 100) +
  scale_y_continuous(limits = c(-10,110)) +
  theme_classic(base_size = 16) +
  theme(strip.background = element_blank())

hrstatus_plot <- ggpar(hrstatus_plot, ylab = bquote(italic('ERBB2')~'CNAP'))

hrstatus_plot

# ggsave(hrstatus_plot, filename = '../../results/2024-03-22_cna-plots/ERBB2_cnap_paired_by_hrstatus.png',
#        width = 5, height = 3, units = 'in', dpi = 450)
```

# Supplementary Figure 2: ERBB2 CNAP by ESR1 mutation status
```{r}
esrmut_plot <- erbb2_tmp_per_pt %>%
  drop_na %>%
  mutate(ESR1_mut = ifelse(participant_id %in% esr1_mut_pts, 'Yes', 'No')) %>%
  pivot_longer(cols = c(Pre, Post)) %>%
  mutate(name = factor(name, levels = c('Pre', 'Post'))) %>%
  ggboxplot(., x = 'ESR1_mut', y = 'value',
            facet.by = 'name',
            line.color = 'gray', line.size = 0.5, add = 'jitter') +
  stat_compare_means(label.y = 100) +
  scale_y_continuous(limits = c(-10,110)) +
  theme_classic(base_size = 16) +
  theme(strip.background = element_blank())

esrmut_plot <- ggpar(esrmut_plot,
                     xlab = bquote(italic('ESR1')~'mutation'),
                     ylab = bquote(italic('ERBB2')~'CNAP'))

esrmut_plot

# ggsave(esrmut_plot, filename = '../../results/2024-03-22_cna-plots/ERBB2_cnap_paired_by_ESR1.png',
#        width = 5, height = 3, units = 'in', dpi = 450)
```

```{r}
sessionInfo()
```
