---
title: "Mutations in key genes, pathway alterations, and signatures - cohort comparisons and summary plots"
output: html_notebook
---

This notebook exists to pull together tabulations and plots of clinical and
genomic variables for the comparisons Kate outlined in a 07/13/2023 email.
```{r}
library(tidyverse)
library(janitor)
library(ggpubr)
library(rstatix)
```

```{r}
output_dir <- '../../results/2024-02-12_pathway-and-signature-plots/'
```

```{r}
patient_sheet_paired <- read_delim('../../results/2024-02-05_cohort-sheets/paired_participant_cohort_sheet.txt', show_col_types = FALSE) %>% 
  mutate(Cohort = 'Paired')

patient_sheet_unpaired <- read_delim('../../results/2024-02-05_cohort-sheets/unpaired_participant_cohort_sheet.txt', show_col_types = FALSE)

patient_sheet <- bind_rows(patient_sheet_paired, patient_sheet_unpaired) %>% 
  mutate(harmonized_tki_duration = ifelse(str_equal(Cohort, 'Paired'), sampled_tki_duration, earliest_tki_duration),
         harmonized_os_tki_dfd = ifelse(str_equal(Cohort, 'Paired'), os_sampled_tki_dfd, os_earliest_tki_dfd))

maf_paired <- read_delim('../../results/2024-02-14_mutations/all_mut_ccfs_maf_bl_geneLists_evidenceLevels.txt',
                         show_col_types = FALSE,
                         col_types = cols(level_of_evidence = col_character())) %>% 
  filter((!str_equal(trajectory, 'low_ccf')) & (!is.na(gene_set)) & (!is.na(level_of_evidence))) %>% 
  distinct(Patient_ID, unique_mut_id, .keep_all = TRUE)

maf_unpaired <- read_delim('../../results/2024-02-12_mutations/all_mut_ccfs_maf_annotated_geneLists_evidenceLevels_unpaired.txt',
                           show_col_types = FALSE,
                           col_types = cols(level_of_evidence = col_character())) %>% 
  filter(postDP_ccf_mode >= 0.1 & (!is.na(gene_set)) & (!is.na(level_of_evidence))) %>% 
  distinct(Patient_ID, unique_mut_id, .keep_all = TRUE)

maf_combined <- bind_rows(maf_paired, maf_unpaired) %>% 
  mutate(gene_set = ifelse(str_equal(gene_set, 'ESR1'), 'ESR1 + regulators', gene_set)) %>% 
  left_join(patient_sheet %>% select(participant_id, Cohort), by = c('Patient_ID' = 'participant_id'))

clone_H <- read_delim('../../merrell/signature_analyses/2023-09-06/clones/H_matrix.tsv',
                        show_col_types = FALSE)

patient_H <- read_delim('../../merrell/signature_analyses/2023-09-06/patients/H_matrix.tsv',
                        show_col_types = FALSE)
```

Paired cohort: # pts with mutations in pathways
```{r}
maf_paired %>% distinct(Patient_ID, gene_set) %>% 
  count(gene_set) %>%
  dplyr::rename(n_pts_mut = n) %>% 
  mutate(pct_pts_mut = round(100 * n_pts_mut / 26)) %>% 
  arrange(desc(pct_pts_mut))
```

Paired vs. pre and paired vs. post cohorts: # pts with mutations in pathways
```{r}
maf_combined %>% distinct(Patient_ID, Cohort, gene_set) %>%
  mutate(Cohort = case_when(
    str_equal(Cohort, 'Exposed') ~ 'Post-TKI',
    str_equal(Cohort, 'Non-exposed') ~ 'Pre-TKI',
    TRUE ~ Cohort)) %>% 
  count(Cohort, gene_set) %>%
  dplyr::rename(n_pts_mut = n) %>% 
  mutate(n_pts_not_mut = case_when(
    str_equal(Cohort, 'Paired') ~ 26 - n_pts_mut,
    str_equal(Cohort, 'Post-TKI') ~ 30 - n_pts_mut,
    str_equal(Cohort, 'Pre-TKI') ~ 55 - n_pts_mut
  )) %>% 
  mutate(pct_pts_mut = round(100 * n_pts_mut / (n_pts_mut + n_pts_not_mut))) %>% 
  arrange(Cohort, desc(pct_pts_mut))
```


Paired vs. pre and paired vs. post cohorts: # pts with mutations in key genes
```{r}
maf_combined %>% distinct(Patient_ID, Cohort, Hugo_Symbol) %>%
  # filter(Hugo_Symbol %in% c('PIK3CA', 'ERBB2', 'ESR1', 'FGFR2', 'FGFR4')) %>% 
  mutate(Cohort = case_when(
    str_equal(Cohort, 'Exposed') ~ 'Post-TKI',
    str_equal(Cohort, 'Non-exposed') ~ 'Pre-TKI',
    TRUE ~ Cohort)) %>% 
  count(Cohort, Hugo_Symbol) %>%
  dplyr::rename(n_pts_mut = n) %>% 
  mutate(n_pts_not_mut = case_when(
    str_equal(Cohort, 'Paired') ~ 26 - n_pts_mut,
    str_equal(Cohort, 'Post-TKI') ~ 30 - n_pts_mut,
    str_equal(Cohort, 'Pre-TKI') ~ 55 - n_pts_mut
  )) %>% 
  mutate(pct_pts_mut = round(100 * n_pts_mut / (n_pts_mut + n_pts_not_mut))) %>% 
  arrange(desc(pct_pts_mut))
```


Paired cohort only: number of participants/pathways mutated for each clonal trajectory
```{r}
labels <- c('truncal' = 'Truncal', 'stable_mixed' = 'Stable', 'growing' = 'Growing', 'shrinking' = 'Shrinking')

n_pathway_per_trajectory <- maf_paired %>%
  group_by(Patient_ID, trajectory) %>%
  summarize(n_pathways = n_distinct(gene_set), .groups = 'drop') %>%
  complete(Patient_ID, trajectory, fill = list(n_pathways = 0)) %>% 
  right_join(patient_sheet_paired %>% select(participant_id, resistance_type),
            by = c('Patient_ID' = 'participant_id')) %>% 
  mutate(trajectory = factor(labels[as.character(trajectory)],
                             levels = c('Truncal', 'Growing', 'Stable', 'Shrinking')),
         resistance_type = factor(resistance_type, levels = c('Intrinsic', 'Acquired')))

n_participants_per_trajectory <- maf_paired %>% 
  left_join(patient_sheet_paired %>% select(participant_id, resistance_type),
            by = c('Patient_ID' = 'participant_id')) %>% 
  group_by(gene_set, trajectory, resistance_type) %>%
  summarize(n_participants = n_distinct(Patient_ID), .groups = 'drop') %>%
  complete(gene_set, trajectory, resistance_type, fill = list(n_participants = 0)) %>% 
  mutate(trajectory = factor(labels[as.character(trajectory)],
                             levels = c('Truncal', 'Growing', 'Stable', 'Shrinking')),
         resistance_type = factor(resistance_type, levels = c('Intrinsic', 'Acquired')))
  
p_pathway_trajectory <- ggboxplot(n_pathway_per_trajectory,
          x = 'resistance_type',
          y = 'n_pathways',
          xlab = 'Resistance type',
          ylab = 'Number of mutated pathways\n(per participant)',
          size = 0.5,
          add = 'dotplot',
          outlier.shape = NA,
          palette = c('#D55E00', '#009E73'),
          facet.by = 'trajectory',
          nrow = 1) +
  theme(legend.position = 'none') +
  stat_compare_means(label = 'p')

p_participant_trajectory <- ggboxplot(n_participants_per_trajectory,
          x = 'resistance_type',
          y = 'n_participants',
          xlab = 'Resistance type',
          ylab = 'Number of participants with\nmutated pathways',
          size = 0.5,
          add = 'dotplot',
          outlier.shape = NA,
          palette = c('#D55E00', '#009E73'),
          facet.by = 'trajectory',
          nrow = 1) +
  theme(legend.position = 'none') +
  stat_compare_means(label = 'p')

ggarrange(p_pathway_trajectory, p_participant_trajectory, nrow = 2) %>%
  ggsave(filename = paste0(output_dir, 'pathway_trajectory_counts.2024-02-14.png'),
         width = 8, height = 6, units = 'in', dpi = 450)
```

Double checking statistical comparisons
```{r}
n_pathway_per_trajectory %>% group_by(trajectory) %>% wilcox_test(n_pathways ~ resistance_type)

n_participants_per_trajectory %>% group_by(trajectory) %>% wilcox_test(n_participants ~ resistance_type)
```

Pathway mutation frequency: acquired, intrinsic, exposed, non-exposed
```{r}
to_plot_muts <- maf_combined %>%
  left_join(patient_sheet, by = c('Patient_ID' = 'participant_id')) %>% 
  mutate(Cohort = factor(case_when(
    str_equal(Cohort, 'Paired') ~ str_c('Paired ', resistance_type),
    str_equal(Cohort, 'Exposed') ~ 'Post-TKI',
    str_equal(Cohort, 'Non-exposed') ~ 'Pre-TKI'
  ), levels = c('Post-TKI', 'Pre-TKI', 'Paired Acquired', 'Paired Intrinsic'))) %>%
  group_by(gene_set, Cohort) %>% 
  summarize(n_pts_mut = n_distinct(Patient_ID), .groups = 'drop') %>%
  complete(gene_set, nesting(Cohort),
           fill = list(n_pts_mut = 0)) %>% 
  mutate(pct_pts_mut = 100 * case_when(
    str_equal(Cohort, 'Paired Acquired') ~ n_pts_mut / 8,
    str_equal(Cohort, 'Paired Intrinsic') ~ n_pts_mut / 18,
    str_equal(Cohort, 'Post-TKI') ~ n_pts_mut / 30,
    str_equal(Cohort, 'Pre-TKI') ~ n_pts_mut / 55
  )) %>% 
  group_by(gene_set) %>% 
  mutate(order_sum = sum(pct_pts_mut)) %>% 
  ungroup %>% 
  arrange(order_sum) %>% 
  mutate(gene_set = replace(gene_set, str_equal(gene_set, 'Cell Cycle'), 'Cell cycle'))

to_plot_muts
```

```{r}
plot_mut_freq <- ggbarplot(to_plot_muts, 'gene_set', 'pct_pts_mut',
          color = 'Cohort', fill = 'Cohort', label = TRUE,
          orientation = 'horiz', lab.nb.digits = 1,
          position = position_dodge(0.75), lab.vjust = 0.5, lab.hjust = -0.1,
          lab.col = 'black',
          ylab = 'Frequency of patients with mutation (%)',
          xlab = 'Pathway',
          lab.size = 4) +
  theme_classic(base_size = 16) +
  theme(legend.position = c(0.8, 0.3)) +
  scale_y_continuous(limits = c(0, 100)) +
  coord_flip(clip = 'off') +
  scale_color_manual(values = c('#0072B2','#F0E442','#009E73','#D55E00')) +
  scale_fill_manual(values = c('#0072B2','#F0E442','#009E73','#D55E00')) +
  guides(color = guide_legend(reverse = TRUE),
         fill = guide_legend(reverse = TRUE))

plot_mut_freq

ggsave(plot_mut_freq, filename = paste0(output_dir, 'mut_freq_pathways.2024-02-14.png'),
       width = 6, height = 8, units = 'in', dpi = 450)
ggsave(plot_mut_freq, filename = paste0(output_dir, 'mut_freq_pathways.2024-02-14.pdf'),
       width = 6, height = 8, units = 'in')
```

Want to test intrinsic vs. acquired and exposed vs. non-exposed with Fisher exact
```{r}
to_fisher <- to_plot_muts %>% 
  select(-pct_pts_mut, -order_sum) %>% 
  mutate(n_pts_not_mut = case_when(
    str_equal(Cohort, 'Paired Acquired') ~ 8 - n_pts_mut,
    str_equal(Cohort, 'Paired Intrinsic') ~ 18 - n_pts_mut,
    str_equal(Cohort, 'Post-TKI') ~ 30 - n_pts_mut,
    str_equal(Cohort, 'Pre-TKI') ~ 55 - n_pts_mut
  ))

fisher_results <- bind_rows(
  to_fisher %>% 
    filter(Cohort %in% c('Paired Intrinsic', 'Paired Acquired')) %>% 
    group_by(gene_set) %>% 
    summarize(data = list(row_wise_fisher_test(
      as.matrix(select(cur_data(), starts_with('n_'))), p.adjust.method = 'BH'))) %>%
    unnest_wider(data) %>% 
    unnest(c(group:p.adj.signif)) %>% 
    mutate(comparison = 'Intrinsic vs. Acquired') %>% 
    distinct(gene_set, p, comparison),
  to_fisher %>% 
    filter(Cohort %in% c('Post-TKI', 'Pre-TKI')) %>% 
    group_by(gene_set) %>% 
    summarize(data = list(row_wise_fisher_test(
      as.matrix(select(cur_data(), starts_with('n_'))), p.adjust.method = 'BH'))) %>%
    unnest_wider(data) %>% 
    unnest(c(group:p.adj.signif)) %>% 
    mutate(comparison = 'Pre-TKI vs. Post-TKI') %>% 
    distinct(gene_set, p, comparison)
)

fisher_results %>% arrange(p)
```

Signature activity: acquired, intrinsic, exposed, non-exposed
```{r}
to_plot_sigs <- patient_H %>% 
  dplyr::rename(`SBS1/5 (clock-like)` = `S1-SBS6`,
         `SBS17b (Capecitabine)` = `S2-SBS17b`,
         `SBS2/13 (APOBEC)` = `S3-SBS2`,
         `SBS26/54 (MMRd/seq. artifact)` = `S4-SBS54`) %>% 
  left_join(patient_sheet, by = c('sample' = 'participant_id')) %>% 
  mutate(Cohort = factor(ifelse(
    str_equal(Cohort, 'Paired'), resistance_type, Cohort
  ), levels = c('Exposed', 'Non-exposed', 'Acquired', 'Intrinsic'))) %>% 
  select(sample, Cohort, contains('SBS')) %>% 
  pivot_longer(cols = contains('SBS')) %>% 
  group_by(sample) %>%
  mutate(value_norm = value / sum(value))

plot_sig_activity <- ggboxplot(to_plot_sigs, x = 'Cohort', y = 'value_norm',
          facet.by = 'name', ylab = 'Signature activity') +
    stat_compare_means(method = 'wilcox.test', label = 'p.signif',
                     comparisons = list(c('Intrinsic', 'Acquired'),
                                        c('Exposed', 'Non-exposed'))) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.25),
                     limits = c(0, 1.25)) +
  theme_bw()

ggsave(plot_sig_activity, filename = paste0(output_dir, 'sig_activity.2024-02-14.png'),
       width = 7, height = 5, units = 'in', dpi = 450)

plot_sig_activity
```
